---
title: "Orthogonal evidence of lncRNA markers in other mCRPC/PCa scRNA-seq datasets"
output: html_notebook
---
#LOAD LIBRARIES
```{r}

library(tidyverse)
library(Seurat)
library(Hmisc)
library(ggrepel)
library(EnhancedVolcano)
library(msigdbr)
library(EnsDb.Hsapiens.v86)
library(Matrix)
library(ggpubr)
library(tidytext)
library(cowplot)
library(ComplexHeatmap)
library(circlize)
library(harmony)
options(ggrepel.max.overlaps = Inf)

```


#READ IN VALIDATION DATA SETS
```{r}
#Read in scRNA-seq data frm He et al. 
bmet <- readRDS(file = "/Users/ds/Desktop/scRNA_datasets/He_scRNA/He_mCRPC.rds")
Idents(bmet) <- "supercluster.for.LDSC.SEG"
final_markers <- read.table(file = "/Users/ds/Desktop/projects/mCRPC_lncRNA_analysis/results/final_markers_mCRPC_He.txt", header = T, sep = "\t")
#Read in scRNA-seq data from Kfoury et al. 
bmcrpc <- readRDS(file = "/Users/ds/Desktop/scRNA_datasets/seurat/Kfoury_bmCRPC_scRNA.rds")
Idents(bmcrpc) <- "cells"
#Read in scRNA-seq data from Song et al. 
scrna <- readRDS(file = "/Users/ds/Desktop/scRNA_datasets/dge_pca.rds")
Idents(scrna) <- "ID_coarse"




```

#FIND OVERLAPPING LNCRNAS ACROSS DATASETS
```{r}
#Intersect datasets
olap <- intersect(row.names(bmcrpc), final_markers$gene) %>% intersect( row.names(scrna))
olap <- dplyr::filter(final_markers, gene %in% olap)$gene



# pdf("/Users/ds/Desktop/plot.pdf", width = 10, height = 10)
# DoHeatmap(object = bmet, features = dplyr::filter(final_markers, gene %in% olap)$gene, slot = "scale.data", )
# dev.off()

```

#PLOT INTERSECTING LNCRNAS IN PSEUDOBULK DATA FROM HE ET AL. 
```{r}

#Return average expression of common lncRNA markers per celltype
res <- do.call(rbind, lapply(X = unique(final_markers$cluster), FUN = function(c){
  
  
  query_lncrna <- final_markers %>%  dplyr::filter(cluster == c & gene %in% olap) %>% arrange(-avg_log2FC)  %>% .$gene
  ae <- AverageExpression(object = bmet, features = query_lncrna, slot = "data") %>%  data.frame() %>% log1p() 
  ae <- ae %>% mutate(Genes = row.names(.), Cluster = c)
  return(ae)

  
} ))

#Format
res <- res %>% mutate(Order = ifelse(Cluster == 'tumor' , 1, ifelse(Cluster == 'B lineage', 2, ifelse(Cluster=='NK/T', 3, ifelse(Cluster == "neutrophil", 4, ifelse(Cluster == "monocyte/macrophage", 5, ifelse(Cluster == "erythroid", 6, 7))  ))))) %>% arrange(Order)
mat <-  dplyr::select(res, -c("Genes", "Cluster", "Order"))
anno <- dplyr::select(res, c("Genes", "Cluster"))
mat <- t(scale(t(mat), scale = T)) 
#mat <- ramify::clip(x = mat, .min = -3, .max = 3)
colnames(mat) <- gsub(pattern = "RNA.", replacement = "", x = colnames(mat))
mat[is.na(mat)] <- 0
anno_col <- colorRampPalette(c("green", "blue", "red", "black", "grey", "purple"))(6)
names(anno_col) <- unique(anno$Cluster)
row_ha = rowAnnotation(Cluster = anno$Cluster, col = list(Cluster = anno_col))


#Plot heatmap
pdf("/Users/ds/Desktop/plot.pdf",  height = 10, width = 7.5)
Heatmap(matrix = mat, right_annotation = row_ha, col = colorRampPalette(c("purple","black", "yellow"))(3), show_column_names = T, cluster_columns = F, na_col = 'black', cluster_rows = F, column_split =as.character(c("A", rep("B", 1),rep("C",1),rep("D", 1), "E" , "F")),row_names_gp = grid::gpar(fontsize = 10) , column_names_gp = gpar(fontsize = 10) ,row_split = c(rep("A",17 ), rep("B", 5), rep("C", 16), rep("D",11 ), rep("E", 8), rep("F", 4)))
dev.off()

mat

```


#PLOT INTERSECTING LNCRNAS IN PSEUDOBULK DATA FROM KFOURY ET AL.
```{r}
#Return average expression of common lncRNA markers per celltype
res <- do.call(rbind, lapply(X = unique(final_markers$cluster), FUN = function(c){
  
  
  query_lncrna <- final_markers %>%  dplyr::filter(cluster == c & gene %in% olap) %>% arrange(-avg_log2FC)  %>% .$gene
  ae <- AverageExpression(object = bmcrpc, features = query_lncrna, slot = "data") %>%  data.frame() %>% log1p() 
  ae <- ae %>% mutate(Genes = row.names(.), Cluster = c)
  return(ae)

  
} ))

#Format
res <- res %>% mutate(Order = ifelse(Cluster == 'tumor' , 1, ifelse(Cluster == 'B lineage', 2, ifelse(Cluster=='NK/T', 3, ifelse(Cluster == "neutrophil", 4, ifelse(Cluster == "monocyte/macrophage", 5, ifelse(Cluster == "erythroid", 6, 7))  ))))) %>% arrange(Order)
mat <-  dplyr::select(res, -c("Genes", "Cluster", "Order"))
anno <- dplyr::select(res, c("Genes", "Cluster"))
mat <- mat %>% dplyr::select(c(paste0("RNA.", c(c("Tumor"),c("Mature.B","memBcell" ,"Pro.B","Immature.B.cells"),  c("CTL.1", "CTL.2", "NK", "NKT", "CD4..Naive" ,"Treg.Active","Th1.17","Treg.Resting","CD8..Naive"),c("Mono3" , "mDC"  ,"TIM" ,"TAM" ,"Mono2"  ,"Mono1","Monocyte.prog"), c("Erythroid")))))#Group similar celltypes together
mat <- t(scale(t(mat), scale = T)) 
mat <- ramify::clip(x = mat, .min = -2, .max = 2)
colnames(mat) <- gsub(pattern = "RNA.", replacement = "", x = colnames(mat))
mat[is.na(mat)] <- 0
anno_col <- colorRampPalette(c("green", "blue", "red", "black", "grey", "purple"))(6)
names(anno_col) <- unique(anno$Cluster)
row_ha = rowAnnotation(Cluster = anno$Cluster, col = list(Cluster = anno_col))


#plot heatmap
pdf("/Users/ds/Desktop/plot1.pdf", width = 10, height = 10)
Heatmap(matrix = mat, right_annotation = row_ha, col = colorRampPalette(c("purple","black", "yellow"))(3), show_column_names = T, cluster_columns = F, na_col = 'black', cluster_rows = F, column_split =as.character(c("A", rep("B", 4),rep("C",9),rep("D", 7), "E" )), row_split = c(rep("A",17 ), rep("B", 5), rep("C", 16), rep("D",11 ), rep("E", 8), rep("F", 4)))
dev.off()



```


#PLOT INTERSECTING LNCRNAS IN PSEUDOBULK DATA FROM SONG ET AL.
```{r}
#Return average expression of common lncRNA markers per celltype
res <- do.call(rbind, lapply(X = unique(final_markers$cluster), FUN = function(c){
  
  
  query_lncrna <- final_markers %>%  dplyr::filter(cluster == c & gene %in% olap) %>% arrange(-avg_log2FC)  %>% .$gene
  ae <- AverageExpression(object = scrna, features = query_lncrna, slot = "data") %>%  data.frame() %>% log1p() 
  ae <- ae %>% mutate(Genes = row.names(.), Cluster = c)
  return(ae)

  
} ))


#Format
res <- res %>% mutate(Order = ifelse(Cluster == 'tumor' , 1, ifelse(Cluster == 'B lineage', 2, ifelse(Cluster=='NK/T', 3, ifelse(Cluster == "neutrophil", 4, ifelse(Cluster == "monocyte/macrophage", 5, ifelse(Cluster == "erythroid", 6, 7))  ))))) %>% arrange(Order)
mat <-  dplyr::select(res, -c("Genes", "Cluster", "Order"))
anno <- dplyr::select(res, c("Genes", "Cluster"))
mat <- mat %>% dplyr::select(c(paste0("RNA.", c(c("Epithelial"),c("B.cell", "Plasma"),  c("T.cell"),c("Myeloid")))))#Group similar cell types
mat <- t(scale(t(mat), scale = T)) 
mat <- ramify::clip(x = mat, .min = -2, .max = 2)
colnames(mat) <- gsub(pattern = "RNA.", replacement = "", x = colnames(mat))
mat[is.na(mat)] <- 0
anno_col <- colorRampPalette(c("green", "blue", "red", "black", "grey", "purple"))(6)
names(anno_col) <- unique(anno$Cluster)
row_ha = rowAnnotation(Cluster = anno$Cluster, col = list(Cluster = anno_col))

#Plot heatmap
pdf("/Users/ds/Desktop/plot2.pdf",  height = 10, width = 7.5)
Heatmap(matrix = mat, right_annotation = row_ha, col = colorRampPalette(c("purple","black", "yellow"))(3), show_column_names = T, cluster_columns = F, na_col = 'black', cluster_rows = F, column_split =as.character(c("A", rep("B", 2),rep("C",1),rep("D", 1))),row_split = c(rep("A",17 ), rep("B", 5), rep("C", 16), rep("D",11 ), rep("E", 8), rep("F", 4)))
dev.off()


```


############Optional validation using CIBERSORTX ##########################

#Create reference for CIBERSORTX GEP estimation (S mode)
```{r}


meta <- meta %>%  mutate(celltype = ifelse(supercluster.for.LDSC.SEG == "NK/T", "NK_T.", ifelse(supercluster.for.LDSC.SEG == "monocyte/macrophage", "monocyte_macrophage.",ifelse(supercluster.for.LDSC.SEG == "B lineage", "B.",ifelse(supercluster.for.LDSC.SEG == "tumor", "tumor.", ifelse(supercluster.for.LDSC.SEG == "erythroid", "erythroid.", supercluster.for.LDSC.SEG)) ) ))) %>% group_by(celltype) %>% mutate(ID = paste0(celltype, 1:n())) 


colnames(tpm) <- meta$ID


reference <- data.frame(Genes = row.names(tpm), tpm)
colnames(reference) <- c("Genes", gsub(x = colnames(reference)[-1], pattern = ".[0-9]*$", replacement = ""))


write.table(reference, file = "/Users/ds/Desktop/He_bulk_reference.txt", quote = F, row.names = F, sep = "\t" )


```

#Plot GEP estimates - see if similar to scRNA data
```{r}
gep <- read.table(file = "/Users/ds/Desktop/projects/lncRNA/results/CIBERSORTxGEP_Job40_GEPs.txt", header = T)
gep <- gep %>% inner_join(y = dplyr::filter(final_markers, gene %in% olap), by = c("GeneSymbol"= "gene"))


row.names(gep) <- gep$GeneSymbol
gep <- gep %>% dplyr::select(c("tumor", "B", "NK_T", "neutrophi", "monocyte_macrophage", "erythroid", "cluster", "GeneSymbol"))

gep
colnames(gep) <- c('tumor', 'B lineage', 'NK/T', 'neutrophil', 'monocyte/macrophage', 'erythroid', 'Cluster', 'GeneSymbol')
gep <- gep %>% mutate(Order = ifelse(Cluster == 'tumor' , 1, ifelse(Cluster == 'B lineage', 2, ifelse(Cluster=='NK/T', 3, ifelse(Cluster == "neutrophil", 4, ifelse(Cluster == "monocyte/macrophage", 5, ifelse(Cluster == "erythroid", 6, 7))  ))))) %>% arrange(Order)


mat <- gep  %>% dplyr::select(-c("GeneSymbol", "Cluster", "Order"))
mat <- mat[olap,]
anno <- gep %>% dplyr::select(c("GeneSymbol", "Cluster"))
anno <- anno[olap,]

mat <- t(scale(t(mat))) %>% data.frame() 
anno_col <- colorRampPalette(c("green", "blue", "red", "black", "grey", "purple"))(6)
names(anno_col) <- unique(anno$Cluster)

row_ha = rowAnnotation(Cluster = anno$Cluster, col = list(Cluster = anno_col))

mat
anno$Cluster

anno$Cluster

pdf("/Users/ds/Desktop/plot.pdf", height = 10, width = 5)
Heatmap(matrix = mat, right_annotation = row_ha, col = colorRampPalette(c("purple","black", "yellow"))(3), show_column_names = T, cluster_rows = F, cluster_columns = F, na_col = 'black', row_split  = c(rep("A",16 ), rep("B", 2), rep("C", 9), rep("D",11 ), rep("E", 3), rep("F", 4)) , column_split = letters[1:6], height = unit(10, 'cm'), row_names_gp = grid::gpar(fontsize = 8) , column_names_gp = gpar(fontsize = 10), cluster_row_slices = F, cluster_column_slices = F)
dev.off()


mat

nrow(mat)/nrow(final_markers)


```

